---
title: "Introduction to lomda: Longitudinal Omics Multivariate Dimension-reduction Analysis"
author: "He Li"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to lomda}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 5,
  out.width  = "100%"
)
```

## Overview

The **lomda** package implements two-stage methods for analysing
high-dimensional longitudinal omics data (metabolomics, proteomics,
transcriptomics, etc.) measured repeatedly over time.

| Stage | Method | Goal |
|-------|--------|------|
| 1     | PCA    | Reduce dimensionality; extract dominant axes of variation |
| 2a    | LMM over `visit` | Model PC trajectories over visit; test for visit effects |
| 2b    | FDA over `age` | Smooth PC trajectories over age; predict age-specific scores |

For the LMM workflow, the model for the $k$-th PC and subject $i$ at visit
$j$ is:

$$
t_{ijk} = \beta_{0k} + \beta_{1k} visit_{ij} + b_{ik} + e_{ijk}.
$$

For the FDA workflow, the PC score is smoothed as a function of age:

$$
t_{ijk} = \mu_k(age_{ij}) + u_{ik}(age_{ij}) + e_{ijk}.
$$

---

## Installation

```r
# From GitHub (development version):
# install.packages("devtools")
devtools::install_github("heli-math/lomda")
```

---

## Quickstart

```{r quickstart}
library(lomda)

# 1. Simulate a longitudinal omics dataset
dat <- simulate_lomda_data(
  n_subjects  = 50,
  n_visits    = 3,
  n_features  = 50,
  n_signal    = 5,     # features with a true time trend
  time_effect = 1,
  seed        = 2026
)

cat("Data dimensions:", nrow(dat), "rows x", ncol(dat), "cols\n")
head(dat[, 1:8])
```

If your metadata columns have different names, pass them explicitly:

```{r custom-columns, eval=FALSE}
fit_custom <- lomda(
  dat,
  ID = "subject_id",
  visit = "wave",
  age = "age_years",
  time = "visit"
)
```

---

## Stage 1: PCA

```{r pca}
# Run PCA + LMM over visit
fit <- lomda(dat, n_pc = 3, time = "visit")
print(fit)
```

### Variance explained

```{r variance, fig.cap="Scree plot showing variance explained by each PC."}
plot(fit, type = "variance")
```

### PC Score Plot (PC1 vs PC2)

```{r scores, fig.cap="PC1 vs PC2 scores coloured by visit."}
plot(fit, type = "scores", color_by = "visit")
```

### Feature Loadings Biplot

```{r loadings, fig.cap="Biplot of feature loadings on PC1 and PC2."}
plot(fit, type = "loadings", n_top = 8)
```

---

## Stage 2: LMM and Inference

### Summary of fixed effects

```{r summary}
summary(fit)
```

### Likelihood Ratio Test

```{r lrt}
lrt_res <- lomda_lrt(fit)
lrt_res
```

### Wald Test (Satterthwaite df via lmerTest)

```{r wald}
wald_res <- lomda_wald(fit, use_t = TRUE)
wald_res
```

---

## Trajectory Plot

```{r trajectory, fig.cap="Mean PC score trajectory over visits."}
plot(fit, type = "trajectory", pcs = 1:3)
```

---

## Stage 2b: FDA Over Age

For age-indexed trajectories, use the same front-door function with
`time = "age"` and choose the FDA smoother with `method_fda`.

```{r fda-age, fig.cap="Age-smoothed trajectory for PC1."}
fit_age <- lomda(dat, n_pc = 3, time = "age", method_fda = "spline")
summary(fit_age)
plot(fit_age, type = "variance")
plot_fda_trajectory(fit_age, pcs = 1:3, n_subjects = 20)
```

```{r fda-predict}
predict(fit_age, ages = c(35, 45, 55, 65))
lomda_important(fit_age, n_top = 10)
```

---

## Advanced: Running Stages Separately

You can run each stage independently:

```{r separate}
# Stage 1
pca_res <- lomda_pca(dat, n_pc = 5)
str(pca_res, max.level = 1)

# Stage 2 (from pre-computed scores)
lmm_res <- lomda_lmm(pca_res$scores, n_pc = 5, adjust = NULL)
names(lmm_res$fits)
```

---

## Session Info

```{r session}
sessionInfo()
```
