---
title: "Introduction to lomda: Longitudinal Omics Multivariate DA"
author: "Your Name"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to lomda}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 5,
  out.width  = "100%"
)
```

## Overview

The **lomda** package implements a two-stage method for analysing
high-dimensional longitudinal omics data (metabolomics, proteomics,
transcriptomics, etc.) measured at repeated clinical visits.

| Stage | Method | Goal |
|-------|--------|------|
| 1     | PCA    | Reduce dimensionality; extract dominant axes of variation |
| 2     | LMM    | Model PC trajectories over time; test for time effects |

The model assumed in Stage 2 for the $k$-th PC and subject $i$ at visit $j$:

$$
\text{PC}_{k,ij} = \beta_0 + \beta_1 t_{ij} + \boldsymbol{\gamma}^\top
\mathbf{z}_{ij} + b_i + \varepsilon_{ij}
$$

where $b_i \sim N(0, \sigma_b^2)$ is a subject-specific random intercept and
$\varepsilon_{ij} \sim N(0, \sigma^2)$.

---

## Installation

```r
# From GitHub (development version):
# install.packages("devtools")
devtools::install_github("yourusername/lomda")
```

---

## Quickstart

```{r quickstart}
library(lomda)

# 1. Simulate a longitudinal omics dataset
dat <- simulate_lomda_data(
  n_subjects  = 60,
  n_visits    = 3,
  n_features  = 50,
  n_signal    = 5,     # features with a true time trend
  time_effect = 0.5,
  seed        = 2024
)

cat("Data dimensions:", nrow(dat), "rows x", ncol(dat), "cols\n")
head(dat[, 1:8])
```

---

## Stage 1: PCA

```{r pca}
# Run both stages together
fit <- lomda(dat, n_pc = 5, covariates = "age")
print(fit)
```

### Variance explained

```{r variance, fig.cap="Scree plot showing variance explained by each PC."}
plot(fit, type = "variance")
```

### PC Score Plot (PC1 vs PC2)

```{r scores, fig.cap="PC1 vs PC2 scores coloured by visit."}
plot(fit, type = "scores", color_by = "time")
```

### Feature Loadings Biplot

```{r loadings, fig.cap="Biplot of feature loadings on PC1 and PC2."}
plot(fit, type = "loadings", n_top = 8)
```

---

## Stage 2: LMM and Inference

### Summary of fixed effects

```{r summary}
summary(fit)
```

### Likelihood Ratio Test

```{r lrt}
lrt_res <- lomda_lrt(fit)
lrt_res
```

### Wald Test (Satterthwaite df via lmerTest)

```{r wald}
wald_res <- lomda_wald(fit, use_t = TRUE)
wald_res
```

---

## Trajectory Plot

```{r trajectory, fig.cap="Mean PC score trajectory over visits."}
plot(fit, type = "trajectory", pcs = 1:3)
```

---

## Advanced: Running Stages Separately

You can run each stage independently:

```{r separate}
# Stage 1
pca_res <- lomda_pca(dat, n_pc = 5, covariates = "age")
str(pca_res, max.level = 1)

# Stage 2 (from pre-computed scores)
lmm_res <- lomda_lmm(pca_res$scores, n_pc = 5, covariates = "age")
names(lmm_res$fits)
```

---

## Session Info

```{r session}
sessionInfo()
```
